### 2010-12-14 docs/tex/conclusion.tex
An incomplete introduction.
Revision updates

%Concluding chapter
\chapter{Conclusion}
If we consider the field of Artificial Life as a tool for biological inquiry then the model presented in this thesis was successful in terms of providing good insight in terms of understanding better the evolutionary process of mimicry as it happens in nature. As mentioned in section \ref{sec:result-conclusion} this model give us proof of the theory of Turner in explaining the evolution of mimicry with punctuated equilibrium. Also it enforces findings of the work of Franks and Noble that multiple Mullerian mimics do not converge into one large ring.

>>>>

%Concluding chapter
\chapter{Conclusion}
\label{chapter:conclusion}
If we consider the field of Artificial Life as a tool for biological inquiry then the model presented in this thesis was successful in terms of providing good insight in terms of understanding better the evolutionary process of mimicry as it happens in nature. As mentioned in section \ref{sec:result-conclusion} this model give us proof of the theory of Turner in explaining the evolution of mimicry with punctuated equilibrium. Also it enforces findings from the work of Frank and Noble that multiple Mullerian mimics do not converge into one large ring.

### 2010-12-14 docs/tex/model.tex
Revision updates

\subsection{Mobility}
An agents position is calculated once during each step of update in time. The agents \(position\), \(force\), \(acceleration\) and \(velocity\) are all vector components. The \(force\) component is calculated from agent's mobility gene. And it is used to compute agent's \(acceleration\). If the \(force\) and \(velocity\) are both zero, then the agent has no effect in motion. Otherwise, Newton's law is used to obtain the \(acceleration\), which is integrated to obtain the new \(velocity\) and new \(position\).

@@

In generating the pattern of Figure \ref{fig:cellular-automata-rule-30}, the genetic representation would be the 'New state of center cell'. As Figure \ref{fig:cellular-automata-rule-30} is a rule 30 cellular automata, the eight bit binary representation of 30 would be the genetic representation for this pattern. Generation of automata pattern from this binary string can be considered analogous to the complex pathway between the genotype and the phenotype that exists in all living organism.

\subsection{Species diversity}
Franks and Noble have used different models to diversify species. In \cite{franks2002} they have used linear difference in number ``constrained to a 'ring' of values  from 1-20 (where 20 and 1 are neighbors)" to distinguish species diversity. Here the distance of one phenotype from another represents their level of similarity.

@@

With CA based pattern representation, population of prey species with a specific CA pattern can be grouped as one single species. And by restricting inter species reproduction we can control the diversity of patterns. But there is mutation applied when similar species mate with each other, so new species do born out of generations of existing species. So that is why we have two separate mutation rate while reproducing prey species. One being the ``Pattern Mutation Rate" (default values are mentioned in table \ref{tab:prey-control-parameters}) with which we control mutation of the first 8 bits of the genome while the ``Genome Mutation Rate" is used to control mutation of rest of the 9 bit genome (table \ref{tab:prey-genome}). Similar efforts of multiple mutation rate at varying location has been used in developing ``Echo" as mentioned in section \ref{subsec:echo} \cite{hraber1997}.

@@

\subsection{Reflection of punctuated equilibrium}
\label{subsec:reflection-of-punctuated-equilibrium}
Punctuated equilibrium has been defined in section \ref{sec:evolutionary-dynamics-of-mimicry} as the evolutionary process that is more inclined to cladogenesis instead of gradualism. Also Turner \cite{turner1988} emphasizes on punctuated equilibrium to describe the evolution of mimicry instead of phyletic gradualism. The design of the model under discussion also follows Turner's explanation in terms of evolving mimicry. As it can be observed new CA patterns evolve from existing ones in prey population just by a single mutation in the pattern gene. Mimics do not follow a gradual process of evolution to look close to models but rather the change happens randomly through a single step mutation. The mutations that are favored, helps the mimics to survive while the unfavored ones fail to persist. It can be observed later in table \ref{tab:diff-in-pattern} how CA patterns of prey species can have vastly different configuration for a unit change in their representative gene.

@@

%This section needs to use similar variables as mentioned in the above section on Hopfield Network.
\subsubsection{Input to memory}
Each prey contains an evolving cellular automata which is represented by a binary gene. This two dimensional pattern is serialized to be available as a one dimensional binary array, which is taken as input for any predator organism trying to interact with the prey. This binary representation of the pattern gets converted to a bipolar representation. Each input pattern will consist of \(\textit{m} \times \textit{n} = \textit{mn}\) components, each component representing one pixel of the pattern (\textit{m} and \textit{n} representing each dimension). The \textit{m} by \textit{n} pattern configuration will be serialized by putting all row vectors in one single row sequentially.

@@

\begin{algorithm}
	\caption{Algorithm for attacking Prey species}
	\label{algo:algorithm-attack-prey}
	\begin{algorithmic}
		\FOR{each step in time}
			\IF {Predator has reached Minimum attack age}
				\STATE $preyToAttack \gets$ select random prey species from the same cell.
				\IF {$currentMemorySize < MinimumMemorySize$}
					\STATE $attemptToKill(preyToAttack);$ \COMMENT {Function defined in Algorithm \ref{algo:algorithm-attemptToKill}}
				\ELSE
					\STATE recognize $preyToAttack$ pattern from memory.
					\IF {prey is not recognized or prey is palatable}
						\STATE $attemptToKill(preyToAttack)$
					\ELSE 
						\STATE Prey is recognized as unpalatable, so release it back into environment.
					\ENDIF
				\ENDIF
			\ENDIF
		\ENDFOR
	\end{algorithmic}
\end{algorithm}

@@

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.60]{images/predators-40}
	\caption[A population of predators]{A population of 40 predators distributed over the environment.}
	\label{fig:predators-40}
\end{figure}

The screen shot in Figure \ref{fig:predators-40} is from the simulation with only 40 predator species. It is presented to show the behavior of predator species in the simulation in absence of any prey species. As it can be observed the predators are distributed all over the cells with a constant mobile behavior to switch to another neighboring cell depending on which ever contains the least number of predator and also most number of prey species. This behavior of predator has been designed to enforce the predatory behavior of these agents and also to have increased predator prey interaction in the simulation in terms of one agent chasing the other for survival of species.

@@

\begin{table}[H]
\centering
\begin{tabular}{| p{2.2cm} | >{\centering} p{2.2cm} | p{8cm} |}
	\hline
		\textbf{Parameter} & \textbf{Value} & \textbf{Description} \\ \hline
		Minimum Memory Size & 2 to 6 & Minimum number of patterns stored in memory before predators start making intelligent decisions.\\ \hline
		Maximum Memory Size & 10 & Maximum number of patterns to be stored in memory. Limited to reduce processing time. \\ \hline 
		Hopfield Maximum Iterations & 20 & Maximum number of iterations for Hopfield Network to recognize a pattern. Usually the network reaches a steady state before that. But this restriction is to avoid infinite loop in case the network never reaches a steady state. \\ \hline
		Attack Age & 500 & minimum age a predator needs to reach to be able to attack prey species.  \\ \hline
		Attack Interval & 100 & Interval of time which needs to pass before a predators attacks its next prey. \\ \hline
		Genome Mutation rate & 0.3 & Mutation rate for the 5 bit genome of the predators representing their mobility and pattern recognition capability. \\ \hline
		Reproduction Age Limit & 500 & Minimum age a predator needs to reach before engaging in reproduction.\\ \hline
		Reproduction Interval & 1000 to 3000 & Interval of time a predator needs to pass between two reproduction process.\\ \hline
		Demise Age & 2000 to 7000 & Age at which a predator is considered as dead.\\
	\hline
\end{tabular}
\caption{Parameters to control predator population and pattern recognition capability.}
\label{tab:predator-control-parameters}
\end{table}

\section{Conclusion}
This model has been designed from a practical point of view; to come up with efficient results and achieve the main objective, \textit{evolution of mimicry}. Creation and transformation of different mimicry ring and also the dynamics of it has been integrated to achieve interesting results. This model can also be considered as a complex adaptive system similar to Hollands work on ``Echo" \cite{holland1996}. The seven basics of a complex adaptive system which are: Aggregation, Tagging, Nonlinearity, Flow, Diversity, Internal Models and Building blocks are present in this model.

>>>>

\subsection{Mobility}
An agent's position is calculated once during each step of update in time. The agents \(position\), \(force\), \(acceleration\) and \(velocity\) are all vector components. The \(force\) component is calculated from agent's mobility gene. And it is used to compute agent's \(acceleration\). If the \(force\) and \(velocity\) are both zero, then the agent has no effect in motion. Otherwise, Newton's law is used to obtain the \(acceleration\), which is integrated to obtain the new \(velocity\) and new \(position\).

@@

In generating the pattern of figure \ref{fig:cellular-automata-rule-30}, the genetic representation would be the `New state of center cell'. As figure \ref{fig:cellular-automata-rule-30} is a rule 30 cellular automata, the eight bit binary representation of 30 would be the genetic representation for this pattern. Generation of automata pattern from this binary string can be considered analogous to the complex pathway between the genotype and the phenotype that exists in all living organism.

\subsection{Species diversity}
Frank and Noble have used different models to diversify species. In \cite{franks2002} they have used linear difference in number ``constrained to a 'ring' of values  from 1-20 (where 20 and 1 are neighbors)" to distinguish species diversity. Here the distance of one phenotype from another represents their level of similarity.

@@

With CA based pattern representation, population of prey species with a specific CA pattern can be grouped as one single species. And by restricting inter species reproduction we can control the diversity of patterns. But there is mutation applied when similar species mate with each other, so new species do born out of generations of existing species. So that is why we have two separate mutation rate while reproducing prey species. One being the ``Pattern Mutation Rate" (default values are mentioned in table \ref{tab:prey-control-parameters}) with which we control mutation of the first 8 bits of the genome while the ``Genome Mutation Rate" is used to control mutation of rest of the 9 bit genome (table \ref{tab:prey-genome}). Similar efforts of multiple mutation rate at varying location has been used in developing Echo as mentioned in section \ref{subsec:echo} \cite{hraber1997}.

@@

\subsection{Reflection of punctuated equilibrium}
\label{subsec:reflection-of-punctuated-equilibrium}
Punctuated equilibrium has been defined in section \ref{sec:evolutionary-dynamics-of-mimicry} as the evolutionary process that is more inclined to cladogenesis instead of gradualism. Also Turner \cite{turner1988} emphasizes on punctuated equilibrium to describe the evolution of mimicry instead of phyletic gradualism. The design of the model under discussion also follows Turner's explanation in terms of evolving mimicry. As it can be observed, new CA patterns evolve from existing ones in prey population just by a single mutation in the pattern gene. Mimics do not follow a gradual process of evolution to look close to models but rather the change happens randomly through a single step mutation. The mutations that are favored, helps the mimics to survive while the unfavored ones fail to persist. It can be observed later in table \ref{tab:diff-in-pattern} how CA patterns of prey species can have vastly different configuration for a unit change in their representative gene.

@@

%This section needs to use similar variables as mentioned in the above section on Hopfield Network.
\subsubsection{Input to memory}
Each prey contains an evolving cellular automata which is represented by a binary genome. This two dimensional pattern is serialized to be available as a one dimensional binary array, which is taken as input for any predator organism trying to interact with the prey. This binary representation of the pattern gets converted to a bipolar representation. Each input pattern consists of \(\textit{m} \times \textit{n} = \textit{mn}\) components, each component representing one pixel of the pattern (\textit{m} and \textit{n} representing each dimension). The \textit{m} by \textit{n} pattern configuration will be serialized by putting all row vectors in one single row sequentially.

@@

\begin{algorithm}
	\caption{Algorithm for attacking Prey species}
	\label{algo:algorithm-attack-prey}
	\begin{algorithmic}
		\FOR{each step in time}
			\IF {Predator has reached Minimum attack age}
				\STATE $preyToAttack \gets$ select random prey species from the same cell.
				\IF {$currentMemorySize < MinimumMemorySize$}
					\STATE $attemptToKill(preyToAttack);$ \COMMENT {Function defined in algorithm \ref{algo:algorithm-attemptToKill}}
				\ELSE
					\STATE recognize $preyToAttack$ pattern from memory.
					\IF {prey is not recognized or prey is palatable}
						\STATE $attemptToKill(preyToAttack)$
					\ELSE 
						\STATE Prey is recognized as unpalatable, so release it back into environment.
					\ENDIF
				\ENDIF
			\ENDIF
		\ENDFOR
	\end{algorithmic}
\end{algorithm}

@@

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.60]{images/predators-40}
	\caption[A population of predators]{A population of 40 predators distributed over the environment.}
	\label{fig:predators-40}
\end{figure}

The screen shot in figure \ref{fig:predators-40} is from the simulation with only 40 predator species. It is presented to show the behavior of predator species in the simulation, in absence of any prey species. As it can be observed the predators are distributed all over the cells with a constant mobile behavior to switch to another neighboring cell depending on the one containing the least number of predator and also most number of prey species. This behavior of predator has been designed to enforce the predatory behavior of these agents and also to have increased predator prey interaction in the simulation in terms of one agent chasing the other for survival of species.

@@

\begin{table}[h!]
\centering
\begin{tabular}{| p{2.2cm} | >{\centering} p{2.2cm} | p{8cm} |}
	\hline
		\textbf{Parameter} & \textbf{Value} & \textbf{Description} \\ \hline
		Minimum Memory Size & 2 to 6 & Minimum number of patterns stored in memory before predators start making intelligent decisions.\\ \hline
		Maximum Memory Size & 10 & Maximum number of patterns to be stored in memory. Limited to reduce processing time. \\ \hline 
		Hopfield Maximum Iterations & 20 & Maximum number of iterations for Hopfield Network to recognize a pattern. Usually the network reaches a steady state before that. But this restriction is to avoid infinite loop in case the network never reaches a steady state. \\ \hline
		Attack Age & 500 & minimum age a predator needs to reach to be able to attack prey species.  \\ \hline
		Attack Interval & 100 & Interval of time which needs to pass before a predators attacks its next prey. \\ \hline
		Genome Mutation rate & 0.3 & Mutation rate for the 5 bit genome of the predators representing their mobility and pattern recognition capability. \\ \hline
		Reproduction Age Limit & 500 & Minimum age a predator needs to reach before engaging in reproduction.\\ \hline
		Reproduction Interval & 1000 to 3000 & Interval of time a predator needs to pass between two reproduction process.\\ \hline
		Demise Age & 2000 to 7000 & Age at which a predator is considered as dead.\\
	\hline
\end{tabular}
\caption{Parameters to control predator population and pattern recognition capability.}
\label{tab:predator-control-parameters}
\end{table}

\section{Conclusion}
This model has been designed from a practical point of view; to come up with efficient results and achieve the main objective, \textit{evolution of mimicry}. Creation and transformation of different mimicry ring and also the dynamics of it has been integrated to achieve interesting results. This model can also be considered as a complex adaptive system similar to Hollands work on Echo \cite{holland1996}. The seven basics of a complex adaptive system which are: Aggregation, Tagging, Nonlinearity, Flow, Diversity, Internal Models and Building blocks are present in this model.
